Supplementary Material for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}

### 2017-10-13 experiments.tex
Correct Appendix references & typos

Figure \ref{fig:all_shd} shows SHDs for all the scoring criteria for each network. Sample sizes range from $10$ to $10000$ and the shown results are averages computed from $1000$ repetitions. None of the scores dominates in all settings considered. BIC fares well when the sample size is low, as it tends to produce nearly empty graph which is a good answer in terms of SHD when the generating networks are relatively sparse. qNML obtains strong results in 'Earthquake' and 'Asia' networks, being the best or the second to best with all the sample sizes considered.

@@

Figure \ref{fig:shd_ranks} summarizes the SHD results in all networks by showing the average rank for each score. The ranking was done by giving the score with the least SHD rank $1$ and the worst one rank $4$. The shown results are averages computed from $5000$ values ($5$ networks, $1000$ repetitions). From this, we can see that qNML has never the worst average ranking, and it has the best ranking with sample sizes greater than $300$. This suggests that qNML is overall a safe choice in structure learning, especially with moderate sample sizes.

@@

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran train and test experiments for
all of them. To better compare the performance over different sample
sizes, we picked different fractions ($10\%, 20\%, \ldots, 90\%$) of
the data sets as training data and the rest as the test data.
This was done for 1000 different permutations of each dataset.
The training was conducted using the dynamic
programming based exact structure learning algorithm.

@@

For each train/test sample, we ranked the predictive performance of
the models learned by different four structure learning criteria (rank
1 being the best and 4 the worst). Table~\ref{tbl:preds} features the
average rank for different data sets, the average being taken over
1000 different train test samples for each 9 sample sizes for each
data.  BIC's bias for simplicity makes it often win (written bold in
the table) with small sample sizes, but it often performs worst
(underlined) for the larger sample sizes (for the same reason), while
fNML seems to be good for large sample sizes. The striking feature
about the qNML is its robustness.  It is usually between BIC and fNML
for all the sample sizes making it a ``safe choice''. This can be
quantified if we further average the columns of the
table~\ref{tbl:preds}.  This yields the average ranks of $2.95, 2.57,
2.10$, and $2.37$, with standard deviations $0.49, 0.90, 0.76$, and
$0.43$.  While fNML achieves on average the best rank, the runner-up
qNML has the lowest standard deviation.

>>>>

Figure \ref{fig:all_shd} shows SHDs for all the scoring criteria for each network. Sample sizes range from $10$ to $10000$ and the shown results are averages computed from $1000$ repetitions. None of the scores dominates in all settings considered. BIC fares well when the sample size is small, as it tends to produce a nearly empty graph which is a good answer in terms of SHD when the generating networks are relatively sparse. qNML obtains strong results in 'Earthquake' and 'Asia' networks, being the best or the second best with all the sample sizes considered.

@@

Figure \ref{fig:shd_ranks} summarizes the SHD results in all networks by showing the average rank for each score. The ranking was done by giving the score with the lowest SHD rank $1$ and the worst one rank $4$. The shown results are averages computed from $5000$ values ($5$ networks, $1000$ repetitions). From this, we can see that qNML never has the worst average ranking, and it has the best ranking with sample sizes greater than $300$. This suggests that qNML is overall a safe choice in structure learning, especially with moderate and large sample sizes.

@@

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran train and test experiments for
all of them. To better compare the performance over different sample
sizes, we picked different fractions ($10\%, 20\%, \ldots, 90\%$) of
the data sets as training data and the rest as the test data.
This was done for 1000 different permutations of each data set.
The training was conducted using the dynamic
programming based exact structure learning algorithm.

@@

For each train/test sample, we ranked the predictive performance of
the models learned by four different structure learning criteria (rank
1 being the best and 4 the worst). Table~\ref{tbl:preds} features the
average rank for different data sets, the average being taken over
1000 different train test samples for each 9 sample sizes for each
data.  BIC's bias for simplicity makes it often win (written bold in
the table) with small sample sizes, but it often performs worst
(underlined) for the larger sample sizes (for the same reason), while
fNML seems to be good for large sample sizes. The striking feature
about the qNML is its robustness.  It is usually between BIC and fNML
for all the sample sizes making it a ``safe choice''. This can be
quantified if we further average the columns of the
table~\ref{tbl:preds}.  This yields the average ranks of $2.95, 2.57,
2.10$, and $2.37$, with standard deviations $0.49, 0.90, 0.76$, and
$0.43$.  While fNML achieves on average the best rank, the runner-up
qNML has the lowest standard deviation.

### 2017-10-13 qnml.tex
fNML result mentioned
Correct Appendix references & typos

We will next state this as a
theorem and sketch a proof for it. A more detailed proof appears in the
supplementary material.
\begin{theorem}
  \label{thm:scoreqv}
  Let $G$ and $G'$ be two Bayesian network structures that differ only
  by a single covered arc reversal, i.e., the arc from $A$ to $B$ in $G$
  has been reversed in $G'$ to point from $B$ to $A$, then
  $$s^{qNML}(D;G)=s^{qNML}(D;G').$$
\end{theorem}
\begin{proof}
  Now the scores for structures can be decomposed as
  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.  Since only the
  terms corresponding to the variables $A$ and $B$ in these sums are
  different, it is enough to show that the sum of these two terms are
  equal for $G$ and $G'$. Since we can assume the data to be fixed we
  lighten up the notation and write
  $P^1_{NML}(i,G_i) := P^1_{NML}(D_{i,G_i};G)$ and
  $P^1_{NML}(G_i)   := P^1_{NML}(D_{G_i};G)$.
  \begin{eqnarray}
    \lefteqn{s_A^{qNML}(D;G)+s_B^{qNML}(D;G)} \nonumber\\
    && =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}
            \frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\nonumber\\
    && =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\nonumber\\
    && =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}
             \frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\nonumber\\
 && =s_A^{qNML}(D;G')+s_B^{qNML}(D;G'),\nonumber
\end{eqnarray}
  using the equations $\{A\}\cup G_A = G_B$, $\{B\}\cup G'_B = G'_A$,
  $\{B\}\cup G_B = \{A\} \cup G'_A$, and $G_A = G'_B$ which follow
  easily from the definition of covered arcs.
\end{proof}

@@

\begin{theorem}
If $G$ consists of $C$ connected components $(G^1,\ldots,G^C)$ with
variable sets $(V^1,\ldots,V^C)$, then $\log P_{NML}(D;G) = s^{qNML}(D;G)$
for all data sets $D$.
\end{theorem}
\begin{proof}
The full proof can be found in Appendix B in the Supplementary
Material.  The proof first shows that NML decomposes for these
particular structures, so it is enough to show the equivalence for
fully connected graphs.
It further derives the number $a(n)$ of
different $n$-node networks whose connected components are
tournaments, which turns out to be the formula for OEIS sequence
A000262\footnote{https://oeis.org/A000108}.
In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051,
37633, 394353, 4596553, \ldots$.
\end{proof}

@@

By analyzing the penalty term of the qNML scoring function, one can prove the following statement:
\begin{theorem}
qNML score is regular.
\end{theorem}
\begin{proof}
The proof is given in the Appendix A in Supplementary Material.
\end{proof}

>>>>

We will next state this as a
theorem and sketch a proof for it. A more detailed proof appears in Appendix A the Supplementary Material.
\begin{theorem}
  \label{thm:scoreqv}
  Let $G$ and $G'$ be two Bayesian network structures that differ only
  by a single covered arc reversal, i.e., the arc from $A$ to $B$ in $G$
  has been reversed in $G'$ to point from $B$ to $A$, then
  $$s^{qNML}(D;G)=s^{qNML}(D;G').$$
\end{theorem}
\begin{proof}
  Now the scores for structures can be decomposed as
  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.  Since only the
  terms corresponding to the variables $A$ and $B$ in these sums are
  different, it is enough to show that the sum of these two terms are
  equal for $G$ and $G'$. Since we can assume the data to be fixed we
  lighten up the notation and write
  $P^1_{NML}(i,G_i) := P^1_{NML}(D_{i,G_i};G)$ and
  $P^1_{NML}(G_i)   := P^1_{NML}(D_{G_i};G)$.
  \begin{eqnarray}
    \lefteqn{s_A^{qNML}(D;G)+s_B^{qNML}(D;G)} \nonumber\\
    && =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}
            \frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\nonumber\\
    && =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\nonumber\\
    && =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}
             \frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\nonumber\\
 && =s_A^{qNML}(D;G')+s_B^{qNML}(D;G'),\nonumber
\end{eqnarray}
  using the equations $\{A\}\cup G_A = G_B$, $\{B\}\cup G'_B = G'_A$,
  $\{B\}\cup G_B = \{A\} \cup G'_A$, and $G_A = G'_B$ which follow
  easily from the definition of covered arcs.
\end{proof}

@@

\begin{theorem}
If $G$ consists of $C$ connected components $(G^1,\ldots,G^C)$ with
variable sets $(V^1,\ldots,V^C)$, then $\log P_{NML}(D;G) = s^{qNML}(D;G)$
for all data sets $D$.
\end{theorem}
\begin{proof}
The full proof can be found in Appendix C in the Supplementary
Material.  The proof first shows that NML decomposes for these
particular structures, so it is enough to show the equivalence for
fully connected graphs.
It further derives the number $a(n)$ of
different $n$-node networks whose connected components are
tournaments, which turns out to be the formula for OEIS sequence
A000262\footnote{https://oeis.org/A000108}.
In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051,
37633, 394353, 4596553, \ldots$.
\end{proof}

@@

By analyzing the penalty term of the qNML scoring function, one can prove the following statement:
\begin{theorem}
qNML score is regular.
\end{theorem}
\begin{proof}
The proof is given in Appendix B in the Supplementary Material.
\end{proof}
